%%% historicity-supplement.tex — Supplementary material
%%% Compile: xelatex historicity-supplement && xelatex historicity-supplement
%%%
%%% Full Greek text (SBLGNT) of the seven undisputed Pauline letters with
%%% every occurrence of the adelph- family of lexemes colour-coded by
%%% classification (A = community, B = biological, C = disputed).

\documentclass[11pt,oneside]{article}

\usepackage[letterpaper,margin=1in,headheight=14pt]{geometry}
\usepackage{fontspec}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{booktabs}
\usepackage{enumitem}
\IfFileExists{csquotes.sty}%
  {\usepackage[autostyle,english=american]{csquotes}}%
  {\providecommand{\enquote}[1]{``##1''}}
\usepackage[colorlinks=true,linkcolor=blue!60!black,urlcolor=blue!60!black]{hyperref}

%% –– Fonts ––
\defaultfontfeatures{Ligatures=TeX}
\setmainfont{STIX Two Text}
\newfontfamily\greekfont{STIX Two Text}
\newfontfamily\greekfontsf{STIX Two Text}

%% –– Headers ––
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\itshape Supplement: \textit{adelph-} occurrences in Paul}
\fancyhead[R]{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}

%% –– Classification colours ––
% A — community / fellow-believer (the overwhelming majority)
\definecolor{colorA}{HTML}{1B7A3E}   % forest green
% B — unambiguously biological sibling (none found in authentic Paul)
\definecolor{colorB}{HTML}{B22222}   % firebrick
% C — disputed or ambiguous
\definecolor{colorC}{HTML}{C77700}   % burnt amber
% D — ethnic / tribal / covenantal kinship
\definecolor{colorD}{HTML}{4A5A9B}   % slate blue
% E — literary / metaphorical extension (not in Paul; defined for cross-doc consistency)
\definecolor{colorE}{HTML}{7E4B9A}   % muted violet

\newcommand{\adeA}[1]{\textcolor{colorA}{\textbf{#1}}}
\newcommand{\adeB}[1]{\textcolor{colorB}{\textbf{#1}}}
\newcommand{\adeC}[1]{\textcolor{colorC}{\textbf{#1}}}
\newcommand{\adeD}[1]{\textcolor{colorD}{\textbf{#1}}}
\newcommand{\adeE}[1]{\textcolor{colorE}{\textbf{#1}}}

%% –– Layout macros for verse/chapter/book ––
\newcommand{\vs}[1]{\textsuperscript{\scriptsize #1}}
\newcommand{\chapheader}[1]{%
  \par\vspace{0.6em}\noindent\textbf{\large #1}\quad\ignorespaces}
\newcommand{\booksection}[2]{%
  \clearpage
  \section*{#1\hfill\normalfont\normalsize\textit{#2}}
  \markboth{#1}{#1}%
  \vspace{0.5em}}

%% –– Fine-tune paragraph formatting for running text ––
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.25em}

\begin{document}

%% –– Title block ––
\begin{center}
  {\LARGE\bfseries Supplementary Material:\\[4pt]
   \textit{Adelph-} Lexeme Occurrences in the\\[2pt]
   Undisputed Pauline Corpus}\\[18pt]
  {\large Companion to \textit{The Evidentiary Status of the Historical Jesus: A Critical Assessment}}\\[6pt]
  {\normalsize Stan Thomas}\\[4pt]
  {\normalsize Independent Researcher, St.\ George, Utah\\
   ORCID: 0000-0002-8828-7856}\\[12pt]
  {\normalsize Draft of \today}
\end{center}

\vspace{1em}

\section*{Purpose}

This supplement reproduces the full Greek text (SBLGNT) of the seven
undisputed Pauline letters---Romans, 1--2 Corinthians, Galatians,
Philippians, 1 Thessalonians, and Philemon---with every occurrence of
the \textit{adelph-} family of lexemes (\textit{adelphos}, G80;
\textit{adelph\=e}, G79; \textit{pseudadelphos}, G5569;
\textit{philadelphia}, G5360) highlighted in colour and classified into
one of three buckets.  Its purpose is to make the lexical argument of
\S3.5 of the main paper (Galatians 1:19 and 1 Corinthians 9:5, the
\textit{adelphos tou kyriou} passages) independently verifiable: every
token is locatable, and the classification of every token is on the
page.

\section*{Classification scheme}

\begin{description}[leftmargin=2em]
\item[\adeA{A — Community / fellow-believer}] The token denotes a member
of the early Christian movement---a \textit{brother in Christ}.  Includes
vocative addresses (``brothers, I urge you\ldots''), named coworkers
identified as ``the brother'' (Timothy, Apollos, Titus, Epaphroditus,
Sosthenes, Quartus, Onesimus), abstract \textit{philadelphia}, and
\textit{pseudadelphos} (``false brother'').  Also includes the
believing-spouse language of 1~Corinthians~7:12--15 (a believer who
happens to be married to an unbeliever), where the term is applied
because of community membership, not marriage.

\item[\adeB{B — Unambiguously biological sibling}] The token denotes a
person's biological brother or sister in a context where no community
reading is defensible.  \textbf{In the authentic Pauline corpus, this
bucket is empty.}  Outside the two disputed ``of the Lord'' passages,
Paul never uses any \textit{adelph-} lexeme to denote a biological
sibling.

\item[\adeC{C — Disputed or ambiguous}] The token admits both a
community and a kinship reading, or neither cleanly.  Three tokens in
the corpus fall here:
\begin{itemize}[leftmargin=2em]
\item \textbf{Rom 16:15} — ``Nereus and his sister''---could be a
biological sibling or simply a fellow-believer.
\item \textbf{1 Cor 9:5} — \textit{hoi adelphoi tou kyriou}, ``the
brothers of the Lord''---plural class label cited parallel to ``the
other apostles'' and ``Cephas.''
\item \textbf{Gal 1:19} — \textit{ton adelphon tou kyriou}, ``James
the brother of the Lord''---the single passage most frequently cited
as decisive for the historicity of Jesus.
\end{itemize}

\item[\adeD{D — Ethnic / tribal / covenantal kinship}] Fellow-Israelite
or fellow-Jew by descent, not by voluntary-community membership.  One
token in Paul:
\begin{itemize}[leftmargin=2em]
\item \textbf{Rom 9:3} — \textit{hyper t\=on adelph\^on mou t\^on
syngen\^on mou kata sarka}, ``on behalf of my brothers, my kinsmen
according to the flesh''---Paul's ethnic kinship with fellow Jews.
The qualifier \textit{kata sarka} marks this as ethnic rather than
community-Christian or nuclear-biological.
\end{itemize}

\item[\adeE{E — Literary / metaphorical extension}] Not used in the
authentic Pauline corpus.  The category is included here for
cross-document consistency with the companion baseline supplement,
where Philo's ``sister-virtues'' and Epictetus's cosmopolitan
``brothers of Zeus'' require this classification.
\end{description}

\section*{Summary tally}

\begin{center}
\begin{tabular}{lrrrrrr}
\toprule
Book & A & B & C & D & E & Total \\
\midrule
Romans           & 18 & 0 & 1 & 1 & 0 & 20 \\
1 Corinthians    & 37 & 0 & 1 & 0 & 0 & 38 \\
2 Corinthians    & 13 & 0 & 0 & 0 & 0 & 13 \\
Galatians        & 11 & 0 & 1 & 0 & 0 & 12 \\
Philippians      &  9 & 0 & 0 & 0 & 0 &  9 \\
1 Thessalonians  & 20 & 0 & 0 & 0 & 0 & 20 \\
Philemon         &  5 & 0 & 0 & 0 & 0 &  5 \\
\midrule
\textbf{Total}   & \textbf{117} & \textbf{0} & \textbf{3} & \textbf{1} & \textbf{0} & \textbf{121} \\
\bottomrule
\end{tabular}
\end{center}

\noindent
121 tokens total (some verses contain two \textit{adelph-} tokens).  The
zero in column B is the central observation: Paul's \textit{adelph-}
vocabulary has no clearly biological-sibling register outside the two
contested ``of the Lord'' cases.  The historicist rebuttal that
\textit{adelphos tou kyriou} is a ``specifically modified kinship
construction'' assumes the existence of a separate biological-sibling
usage from which the modified form is drawn; no such usage exists in
the authentic corpus.

\vspace{0.5em}
\noindent\textbf{Colour legend:} \adeA{A (community)} —
\adeB{B (biological)} — \adeC{C (disputed)} — \adeD{D (ethnic)} —
\adeE{E (metaphorical)}.

\section*{Source and method}

The Greek text is the \textbf{SBL Greek New Testament} (SBLGNT, 2010),
edited by Michael W.\ Holmes and published by the Society of Biblical
Literature.  It is distributed under a Creative Commons
Attribution~4.0 International licence.  The text has been reproduced
here with editorial sigla (diamonds, brackets, and dots marking
apparatus variants) silently removed for readability; no words have
been altered.

Tokens were located by regular-expression matching of any Greek word
containing the root substring \textit{λφ} and cross-verified against
the Englishman's Greek Concordance on Bible Hub.  Verse-by-verse
classifications were assigned by the author; the machinery that
converts classifications into colour markup is a short Python script
included in the project repository at
\url{https://github.com/sbthomas/historicity}.

\vspace{0.5em}
\noindent\textbf{Colour legend:} \adeA{green bold = A (community)} —
\adeB{red bold = B (biological)} —
\adeC{amber bold = C (disputed)} —
\adeD{blue bold = D (ethnic)} —
\adeE{violet bold = E (metaphorical)}.

\clearpage

%% ============================================================
%%  The full running text
%% ============================================================

\input{supplement_body.tex}

\clearpage

%% ============================================================
%%  Appendix: the non-A tokens in context (C and D)
%% ============================================================

\section*{Appendix: The non-A-classified tokens in context}

\subsection*{Romans 9:3 \textbf{(D)} --- \textit{adelph\^on m\=ou t\^on syngen\^on m\=ou kata sarka}}

\noindent\enquote{I could wish myself accursed from Christ on behalf
of my \adeD{brothers}, my kinsmen \textit{according to the flesh}.}

Paul's qualifier \textit{kata sarka} (``according to the flesh'')
marks this use as ethnic/tribal rather than nuclear-family kinship:
the ``brothers'' in view are fellow Israelites, the Jewish people as
a collective.  This is neither community-Christian usage (which would
not require the qualifier) nor a reference to biological siblings.
Classified D under the A/B/C/D/E scheme shared with the baseline
supplement.  Does not support a biological-sibling reading of
\textit{adelphos} elsewhere in Paul.

\subsection*{Romans 16:15 --- \textit{N\=erea kai t\=en adelph\=en autou}}

\noindent\enquote{Greet Philologus and Julia, Nereus and his
\adeC{sister}, and Olympas, and all the saints who are with them.}

The pronoun \textit{autou} (``his'') makes biological kinship
grammatically natural: a greeting to Nereus and the sister he happens
to have.  But it is equally consistent with a fellow-believer who is
in a ``sister of''-style relation to Nereus within the community
(compare Phoebe described as \textit{h\=e adelph\=e h\=em\=on},
``our sister,'' in 16:1).  Classified C to be conservative; the
biological reading is defensible here, but not forced.

\subsection*{1 Corinthians 9:5 --- \textit{hoi adelphoi tou kyriou}}

\noindent\enquote{Do we not have the right to be accompanied by a
believing wife [\textit{adelph\=en gynaika}, classified A], as do the
other apostles and \adeC{the brothers of the Lord} and Cephas?}

Paul enumerates three categories of authority: (1) the other
apostles, (2) the brothers of the Lord, (3) Cephas.  The phrase
\textit{hoi adelphoi tou kyriou} is plural and functions as a
class-label, not a biographical identifier.  If this phrase means
``biological siblings of Jesus,'' the entire collective of Jesus's
siblings is travelling the Mediterranean as recognised authority
figures, cited parallel to the apostolic college and Peter
individually.

\subsection*{Galatians 1:19 --- \textit{t\^on adelph\^on tou kyriou}}

\noindent\enquote{Of the other apostles I saw none, except James,
\adeC{the brother of the Lord}.}

The single passage most frequently cited as decisive for historicity.
The biological reading (James as Jesus's biological sibling) is
possible; the community-title reading (James as a member of a
recognised group called ``brothers of the Lord,'' see 1~Cor~9:5) is
equally compatible with Paul's phrasing and more parsimonious with
his usage elsewhere.  Paul's default meaning of \textit{adelphos}
throughout the corpus is fellow-believer (117 A-classified
occurrences), and the fixed referent of \textit{ho kyrios} in
Paul---the risen, cosmic Christ, not a recently-living
Galilean---tells against the biological reading.  The construction
\textit{adelphos tou kyriou} does not, in Paul's vocabulary, have a
clear biological template from which a kinship reading could be
drawn.

\vspace{2em}

\noindent\rule{0.5\textwidth}{0.4pt}\par
\vspace{0.5em}
\small\noindent\textit{Attribution:} Greek text of the New Testament
from the SBL Greek New Testament, edited by Michael W.\ Holmes.
Copyright~\textcopyright~2010, Society of Biblical Literature and
Logos Bible Software.  Used under a Creative Commons Attribution~4.0
International licence (\url{https://creativecommons.org/licenses/by/4.0/}).

\end{document}
